---
title: "Exploring Health Insurance Costs & Factors"
format: pdf
editor: source
author: "Acree Tejada"
---

```{r}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(moderndive)
library(gridExtra)
library(mosaic)
library(corrplot)
library(GGally)
library(emmeans)
library(ggResidpanel)
library(ggformula)
library(skimr)
library(kableExtra)
library(dplyr)


# to install and load additional packages, remove # and run following
# install.packages("package_name")
# library(package_name)
```



## Introduction

In this study, I will be going over potential factors that influence cost of health 
insurance. Specifically, exploring two potential factors, region location and the number 
of children an individual has as predictors of their medical expenses. Understanding
which factors drive health insurance costs is beneficial for individuals who are 
selecting or planning changes to their coverage, moreso when considering having 
children or relocating. With knowledge, consumers can make informed financial decisions
and manage future expenses. 

Based on background knowledge, this leads to some hypothesis: 
1) Region and price of health insurance does not have any relationship
2) The number of children has no correlation with price of health insurance
3) The effect of the number of children individuals have on the price of health 
insurance does not depend on the region they live in.



## Data

```{r}
#| echo: false
#| message: false
#| warning: false

insurance <- read_csv("insurance.csv") 
# code for the six optional datasets are given below, remove # from the one you will use
# for any other dataset, consult Prof. Chakraborty (the code should look similar if you obtain a .csv file)

# bikes <- read_csv("bikes.csv") 
# flights <- read_csv("flights.csv") 
#insurance <- read_csv("insurance.csv") 
# legos <- read_csv("legos.csv") 
# movies <- read_csv("movies.csv") 
# soccer <- read_csv("soccer.csv") 


```


```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| fig-align: 'center'
#| message: false
#| warning: false

ggplot(data = insurance) +
  geom_boxplot(aes(x = region, y = charges)) +
  labs(x = "Living space by region", y = "Price of Health Insurance Costs (in thousands)", title = "Health Insurance costs by Region")

```

This graph looks at region as an explanatory variable to see whether location impacts
the price of health insurance. From the boxplot, we can see that all four regions 
show very similar outcomes, with most median insurance costs falling roughly between
8,000 and 10,000. This suggests that location alone does not drastically change 
typical insurance prices. For someone considering moving to a new region, this is
useful because it shows that insurance cost is not heavily tied to where you live.

However, the plot also shows a noticeable spread within each region, meaning there
are other important factors influencing insurance cost beyond location. The minimum
values across regions fall around 1,000–2,000, while the 25th percentile is near 5,000.
The 75th percentile varies more, with the southwest appearing slightly lower and the
southeast slightly higher. The same pattern holds for maximum values. Overall, while
a slight difference exists such as the southwest trending lower and the southeast
trending higher, the similarities across regions suggest that location is not the main
influence of health insurance cost.


```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
#| fig-align: 'center'
#| message: false
#| warning: false


ggplot(data = insurance) +
  geom_point(aes(x = children, y = charges)) +
  facet_wrap(~ region) +
  labs(x = "Number of Children", y = "Price of Health Insurance Costs (in thousands)", color = "Number of Children", title = "Health Insurance costs by Region and Number of Children")

```

This graph examines how both region and number of children relate to health insurance
cost. Each panel represents one region, allowing us to compare patterns across locations.
Overall, we can see that as the number of children increases, insurance costs tend to
remain similar with some decreases in every region, although the amount of variation
within is a bit spread. The southeast appears to show slightly higher insurance costs
on average, while the southwest generally appears somewhat lower, but the differences
between regions are not dramatic. For families with four or more children, the sample
size becomes much smaller, so any conclusions for those groups should be interpreted
cautiously. Overall, this plot suggests that the number of children has a more noticeable
effect on insurance cost than region does, but both variables contribute in some way.


```{r}
children <- insurance |> select_if(is.numeric)

corrplot(cor(children))

```

Based on the correlation plot, the correlations between charges and number of
children appear pretty weak, indicating that children count does not show 
a strong linear association with health insurance costs. Most values on this plot
are close to zero, supporting the idea that insurance charges may depend on other 
factors not included in this study.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: 'center'
#| message: false
#| warning: false

table_summary <- insurance |>
  group_by(region, children) |>
  summarize(mean_cost = mean(charges,na.rm=TRUE),
            median_cost = median(charges, na.rm=TRUE),
            sd_cost = sd(charges, na.rm=TRUE),
            n = n())

knitr::kable(table_summary, caption = "Summary of Health Insurance Costs by Region and Number of Children")


```

The table above summarizes health insurance costs by both region and number of children,
displaying mean, median, standard deviation, and sample size for each group. Overall,
the average insurance cost fluctuates up and down as the number of children increases
across all regions, supporting the trend observed in the scatterplot. For example, 
in both the northeast and southeast regions, the mean cost fluctuates from families 
with zero children to families with three children, and then decreases drastically 
when looking at families with four or five children. While there are small regional 
differences, they are not dramatic; the southeast consistently shows slightly higher 
average costs, while the southwest tends to show slightly lower averages. The standard 
deviation values across groups are quite large, showing that insurance costs vary 
widely within each region and child count category. This high variability suggests 
that, beyond location and number of children, other factors such as age, BMI, and 
smoking status likely play a major role in determining insurance cost. In general, 
the table reinforces the idea that the number of children has a stronger and more
consistent association with insurance costs than region alone, while still showing
variation across each region.


## Models

For the 1st model I chose region as the explanatory variable because it's known 
that health insurance costs tends to vary from place to place, and it seemed
reasonable to explore whether region alone contributes to difference in charges.
Also the way financial systems and pricing works is different across locations so
assuming same cost across all regions would not be appropriate. The mathematical 
equation for the model is:
$$
\widehat{\text{charges}} = \beta_0 + \beta_1\times\text{regionNW} + 
\beta_2\times\text{regionSE} + \beta_3\times\text{regionSW}.
$$
I fitted the model in R using the lm() function, then reviewed the summary output.
Upon examining the assumptions, the residual plot showed no curve, suggesting 
linearity is acceptable, but the vertical grouping shows potential violation in 
constant variance. The QQ plot showed a great amount of divergence from the diagonal
line on both ends, violating normality assumption. The index plot showed random scatter,
satisfying the independence, and the histogram showed right-skewedness, meaning the
model tends to underpredict high charges. I didn't include any transformations, 
nonlinear terms, or interactions in this model because I wanted to see the effect
of region alone towards insurance costs.


```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: 'center'

model1 <- lm(charges ~ region, data = insurance)
summary(model1)

resid_panel(model1, smoother = TRUE)

```


For the 2nd model, I added number of children as an explanatory variable to see if
this factor could better explain the variation in insurance costs. Additionally, since 
Log transformation is known to help with normality violations (right-skewed data), 
I proceeded to apply it to model 2 so it can address that concern. The equation for
the model is:
$$
\text{log(charges)} = \beta_0 + \beta_1\times\text{regionNW} + \beta_2\times\text{children} 
+ \beta_3\times\text{regionSE} + \beta_4\times\text{regionSW}
$$
I fitted this model in R, where the log transformation improved normality greatly
as seen in the symmetry/bell-shape from the histogram and the closer points to the 
diagonal line on the QQ plot. However, there is still a slight funnel shape 
(heteroscedasticity), and the residuals for the children plot showed a clear non-linear
trend at the far right, giving us the assumption of constant variance and linearity
still being a concern.


```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: 'center'

model2 <- lm(log(charges) ~ region + children, data = insurance)
summary(model2)

resid_panel(model2, smoother = TRUE)
resid_xpanel(model2, smoother = TRUE)
exp(coef(model2))
exp(confint(model2, "children", level = 0.95))

```


For the 3rd model I kept region and number of children, log-transformation, and 
added interaction between region and number of children to see if the effect
of having more children on insurance costs depended on the specific region.
The equation for model 3 is:
$$
\text{log(charges)} = \beta_0 + \beta_1\times\text{regionNW}
+ \beta_2\times\text{regionSE} + \beta_3\times\text{regionSW} +
+ \beta_4\times\text{children} +
\beta_5\times(\text{regionNW}\times\text{children}) + 
\beta_6\times(\text{regionSE}\times\text{children}) + 
\beta_7\times(\text{regionSW}\times\text{children})
$$
I again fitted the model in R, where the assumption checks for model 3 were nearly
identical to those of model 2, while also still showing good normality but some 
concerns about constant variance and linearity.


```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: 'center'

model3 <- lm(log(charges) ~ region * children, data = insurance)
summary(model3)

resid_panel(model3, smoother = TRUE)

```


For the 4th model I decided to keep region and children, take out the interaction
between the two, and instead add a quadratic term for children to resolve the linearity
assumption violation. The equation for model 4 is:
$$
\text{log(charges)} = \beta_0 + \beta_1\times\text{regionNW} +
\beta_2\times\text{regionSE} + \beta_3\times\text{regionSW} +
\beta_4\times\text{children} + \beta_5\times\text{children}^2
$$
I fitted this model in R, where now if we look at the assumptions, each remain the
same, besides the residual plot; it now doesn't violate the linearity assumption, 
but still violates the constant variance assumption.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: 'center'

model4 <- lm(log(charges) ~ region + children + I(children^2), data = insurance)
summary(model4)

resid_panel(model4, smoother = TRUE)
resid_xpanel(model4, smoother = TRUE)
exp(coef(model4))

exp(confint(model4, "I(children^2)", level = 0.95))

```

## Results

Firstly, it was rather surprising finding out that Region as an explanatory variable
barely explains the variability in price of health insurance. Looking at model 1, 
based on the summary of R, we obtained an $R^2$ of 0.006, which is roughly 0.6% of 
the variability in price of health insurance. Not only that but we can see 
that each region has a p-value over 0.05, meaning that there's no strong evidence 
that region alone has an effect on insurance price. Additionally our assumptions 
showed constant variance and normality violations so adding log transformation 
was part of the next step.

In model 2, it was interesting that applying log transformation and number of children
as an explanatory variable in addition to region, that our $R^2$ = 0.026, 
which is 2.6%. The number of children became significant (p-value approximately 0),
telling us that families with more children tend to have higher insurance costs
(21% increase). While most region's had a p-value over 0.05, southwest this time had
a p-value of 0.04, showing significantly lower average insurance costs than the northeast
region, telling us that taking number of children into account helped explain additional
variability, a more accurate estimate. Based on the 95% confidence interval, we see 
that for each additional child, health insurance cost is estimated to increase by
approximately 7% to 18%, as the multiplicative effect is 1.07 to 1.18. Adding log
transformation improved normality, but some heteroscedasticity remained, probably 
because of the uneven spread across predicted values and the small number of observations
for families with four or five children.

In model 3, after adding an interaction term we see that the p-value of 
northwest region alone lowered to 0.05, all others increased above 0.05, and as 
for children, the p-value increased, but remained at 0.05. Additionally, each interaction 
between region and children had a p-value above 0.05, and our $R^2$ value increased very 
slightly to 0.028, a 2.8% compared to model 2's 2.6%. This tells us that the interaction
term is not significant, providing no strong evidence that the effect of children on
insurance costs differs by region. It only added complexity, therefore model 2 is a
better fit.

Lastly, in mode 4, after adding a quadratic term to children, we see that linearity
is addressed and no longer violated. Not only that, but if we look at the p-value for
children^2, it is less than 0.05, meaning the quadratic term is very significant.
Furthermore, if we look at our $R^2$ value, it now increased to 0.03429, which is
3.4%, which is an even bigger increase compared to model 2's $R^2$ value of 0.026 (2.6%).
Diving deeper, the positive linear term for children and negative quadratic term for
children^2 indicates a non-constant, non-linear relationship. This tells us that
insurance increases with the first few children, but the rate of increase slows down
as the number of children gets higher. Also, the southwest region remains significant
(p = 0.04) and is associated with lower health insurance costs compared to the northeast
baseline, while the northwest and southeast regions show no significant difference.
Based on the 95% confidence interval, we are 95% confident that for every one-unit increase
in children^2, the average health insurance cost is multiplied by a factor between 
0.921 (7.9% decrease in the rate of change) and 0.986 (1.4% decrease in the rate of change).
Overall, because model 4 addresses the non-linearity assumption and gives a higher
$R^2$, it is deemed the better model out of the four for this analysis. 


## Conclusion

Overall, it seems that Region and Number of Children barely explained any variability
in price of health insurance. Although the number of children showed a statistically 
significant positive relationship with insurance cost, the best-fitting Model 4 still 
had a weak overall fit, showing only 3.4% of variability. Furthermore, we found no 
meaningful interaction between region and children, supporting the hypothesis that 
the effect of family size does not depend on location. Crucially, Model 4 revealed 
that the relationship between children and cost is non-linear: costs increase with 
the first few children, but the rate of increase diminishes as the number of children rises.

For medical professionals and policymakers, this study highlights that family size and 
regional differences tend to be poor predictors of health insurance costs. The low $R^2$ 
values tell us that there are other unexamined factors such as age, sex, BMI, and smoking 
status that potentially explain more variability in determining health insurance prices. 
All things considered, our conclusions are limited by the small sample sizes for families 
with four or five children and the unresolved concern regarding the constant variance 
assumption in the models. Specifically, the constant variance violation in the residual
plot means the standard errors and p-values reported may be less reliable. To create
more precise or accurate models/fairer policies, stakeholders should consider other
additional factors as those are more likely to explain a larger portion of cost variation.


## Citations/References

ChatGPT (word structuring/clarification). 

https://www.kaggle.com/datasets/mirichoi0218/insurance 

